From physical principles to ???

### 2016-01-20 2015CompleteDerivation/completeDerivation.tex
More on state variable evolution

To keep track of the state evolution, we'll need to understand how state variables change under continuous and reversible evolution. The first approach is to study the change in the state variables values from $q^i(\mathcal{s})$ to $q^i(\hat{\mathcal{s}})$. This is especially useful when state variables correspond to physically meaningful quantities. For example, we track how the temperature or the pressure of an ideal gas has changed. However, what we'll want to do is to find relationships that are common for all deterministic and reversible processes, which is hard to do in this picture. The second approach is to evolve the state variables as well. The idea is to keep the connection to the initial state by labelling the future state, instead by the future values of $q^i$, by the original one. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$. We can always do this because we have bijective correspondence between initial and final states. This makes it easier to study relationships that are invariant during the evolution: once we verify them for $q^i$ on the initial states, they will automatically hold for the $\hat{q}^i$ at later (or previous) times. Naturally, as $\hat{q}^i$ are new variables that will not hold the same physical meaning. For example, each will be a combination of pressure and temperature that depends on the particular process. We'll typically use the first picture to describe equations of motions and the second to describe invariants. Note, though, that the two picture are tightly linked. In fact:
\begin{align*}
\Delta q^i &= q^i(\hat{\mathcal{s}}) - q^i(\mathcal{s}) \\
\hat{q}^i(\hat{\mathcal{s}}) &= q^i(\hat{\mathcal{s}}) - \Delta q^i
\end{align*}

@@

When assembling multiple state variable into a state vector, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen item, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have to value. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=s$ is available; for $n=2$ we can choose $l=s$ or $l=p$. The choices are not independent. That is: two or more variables are independent if there always exist a state for any possible combination, if the total number of states is the cartesian product of the possibilities of each variable. For discrete space variables,

@@

TODO: labels/state variables, cardinality of state variable, state id function (=1 for a particular state). Can we do for continuous variables as well? Can we find divergence free evolution for det/rev and independent variables?

>>>>

To study time evolution, we need to describe how state variables change under continuous and reversible evolution. There are two ways to do it, and we'll need both. The first approach is to \emph{evolve} the state variables from the initial value $q^i(\mathcal{s})$ to the final value $q^i(\hat{\mathcal{s}})$. The result is a trajectory $q^i(t)$ which is especially useful when state variables correspond to physically meaningful quantities. For example, we track how the temperature or the pressure of an ideal gas changes. Evolved state variables, however, make it hard to find relationships that are invariant and common to all deterministic and reversible processes.

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labelling the future state, instead by the future values of $q^i$, by the original one. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The level sets of $\hat{q}^i$ are the evolved level sets of $q^i$, which allows us to study how groups of states evolve in time. Moreover, once we identify an invariant for $q^i$ this is automatically satisfied by $\hat{q}^i$. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variable provide a way to label the trajectories themselves. While the numeric value of the transported state variables remains the same, the state variable (and its physical meaning) changes. For example, pressure and temperature will be combined depending on the particular process.

Evolved state variables are useful to write equations of motions, study how physical quantities change, form a physical picture of what happens. Transported state variables are useful to write invariants, study state space trajectories, from a geometric picture for the state space. The two, though, are tightly linked. In fact:
\begin{align*}
\Delta q^i &= q^i(\hat{\mathcal{s}}) - q^i(\mathcal{s}) \\
\hat{q}^i(\hat{\mathcal{s}}) &= q^i(\hat{\mathcal{s}}) - \Delta q^i
\end{align*}
Intuitively, the transported variables change in the oppose way as the evolution, so that the value of the variables remain the same.\footnote{One should not confuse evolved/transported state variable with active/passive transformations or with Schroedinger/Heisenberg pictures. In those cases, the choice is whether to change the state or the coordinates/observables. In our case, the state is always changing. The choice is whether to change the state variables or their value.}

@@

\begin{prop}\label{transported variable}
TODO: Define and prove existence
\end{prop}

When assembling multiple state variables into a state vector, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen item, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have two values. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=s$ is available; for $n=2$ we can choose $l=s$ or $l=p$. The choices are not independent. That is: two or more variables are independent if there always exists a state for any possible combination, if the total number of states is the cartesian product of the possibilities of each variable.

For discrete state variables, the total number of states is the product of the possibilities of each independent state variable. That is: if $\Delta q^1$ and $Delta q^2$ are range of possibilities for two independent variables, the total number of possibilities $\#(\Delta q^1) \#(\Delta q^2)$. These relationships are maintained during evolution by the transported variables. Since the value remains unchanged, the range of possibilities for each value remain the same: $\#(\Delta \hat{q}^i) = \#(\Delta q^i)$. The variables remain independent as well, all possible combination remain, and we have: $\#(\Delta \hat{q}^1) \#(\Delta \hat{q}^2) = \#(\Delta q^1) \#(\Delta q^2)$. These relationship will later be extended in the continuous case to independent degrees of freedom for a Hamiltonian system, and will justify the symplectic form. Though this can't be done in general and requires further assumptions.

In the case of continuous state, if each state can be identified by n independent state variables then, at least locally, we can chart the state space on $\mathcal{R}$: the state space is a manifold. Studying in general how ranges of possibilities evolve, though, becomes problematic. Time evolution will map points, while range of continuous possibilities will be length. We would need a metric, invariant during the evolution, telling us how ranges of possibilities map from one variable to another, possibly describing different physical quantities. We will not study this case, but we do note that, if such a metric exist, it depends on the system and on the choice of state variables.

@@

We have seen that physical systems undergoing deterministic and reversible evolution are properly described by a dynamical system. Systems described by a discrete set of states have nice properties for counting possibilities of independent variables, which are maintained during the evolution. System described by a continuous set of states, instead, do not posses in general such properties.
